---
title: "Final Project Stat 5310"
author: "Otavek Aliev | Oscar Pacheco"
date: "`r Sys.Date()`"
output: pdf_document
--- 
# Introduction
Several psychologists agree that the main goal of rational human beings is to find 
happiness irrespective of their geographical locations, the time they lived or other 
demographic variables. Happiness is defined differently in each community, hence, 
there is no common ground to describe happiness (McMahon, 2008). However, there are 
entities that go deep in trying to answer and assess the happiness of the world, 
by taking worldwide surveys in all the countries, and utilizing distinct factors, 
to measure happiness. 

For our project, we will use the World Happiness Reports from the year 2021, 
provided by the Gallup World Survey year 2021, which is a poll that continually 
surveys citizens in 160 countries, representing more than 98% of the world's adult 
population. The poll has over 100 global questions and region-specific items. 
Gallup also works with organizations, cities, governments, and countries to create 
custom items and indexes to gather information to provide a score to the “happiness” 
of each country. 

The main purpose of our regression analysis is to predict the world Ladder.score - 
Happiness score (Dependent variable) based on different variables like GDP per 
capita, social support, healthy life expectancy, freedom to make life choices, 
generosity (independent variables). We will also determine which variables have 
more influence on the response variable. Lastly, we hope to provide a distinct 
perspective to our readers, as often, we define happiness with material things, 
and we do not measure it with other variables that are important in our lives, 
like peace, freedom of choices, generosity, etc. 

## Approach for Regression analysis 

We will start with data exploration by viewing the basic statistics of the variables , 
different data types, and review if there is need for data preprocessing. We then 
will proceed to visualize the data, plotting the different variables and see the 
correlation with a scatterplot matrix. 

We will develop our regression models based on the most relevant features for our 
dependent variable. Run different models to pick the most accurate one and 
improve the model. Finally, we will show our prediction based on the selected best 
fitted model.
# Library, packages and data 

```{r, results =FALSE, message=FALSE}
library(ggplot2) 
library(tidyverse) 
library(GGally) 
library(corrplot) 
library(compareDF)

hr_2021 <- read.csv("2021.csv", stringsAsFactors = TRUE) #World Happiness Report 2021
```

# Data Preparation
## Data Source 
Dataset “World Happiness Report 2021”, was obtained on Kaggle, the link is attached below: 

https://www.kaggle.com/datasets/mathurinache/world-happiness-report?select=2020.csv

The data set “The World Happiness Report”, is a publication of the Network, 
powered by the Gallup World Poll data.

## Dataset 
The World Happiness Report 2021 contains 149 observations and 20. It contains 18 
numerical variables like "Ladder.score"(happiness score), "Logged.GDP.per.capita",
"Generosity" among other scores. Also contains 2 nominal variables: "Country" and 
"Regional.indicator"(Most useful for our analysis).

Top 3 countries: Finland (1) | Denmark (2) | Switzerland (2)

Bottom 3 countries: Rwanda (147) | Zimbabwe (148) | Afghanistan (149)

Rank of the Unites States: United States (19)

```{r , message=FALSE}
#Top 3 countries
head(hr_2021, n = 3)
#Bottom 3 countries
tail(hr_2021, n = 3)
#United States Rank
hr_2021[which(hr_2021$Country.name == "United States"), ]
```
# Data Variables

## Nomival variables
We can see that “Regional.indicator” is not distributed in a proportional way, 
since there are regions in the world with more countries than others. 

```{r}
table(hr_2021$Regional.indicator)
```

## Numerical variables 
Our dependent variable (Ladder.score) shows that the scores range from 2 - 7 approx., 
in a scale from 0 - 10, with a median of 5 and a mean also of 5, showing how most 
of the country's "happiness" is. The summary of the data set also allows us to see 
that some variables are whole numbers of scores and some are percentages such as
"perceptions.of.corruptions" and  "Generosity". 

```{r, }
summary(hr_2021)
```

## Data Cleaning
We have no missing values in our data set, and with "str" function, we can check 
if there are any inconsistent factors, which we do not have in our data set.
We also were able to see some variables have data that's repetitive so we will 
make them "NULL", data ready for exploratory analysis. 

```{r, message=FALSE}
sum(is.na(hr_2021))
str(hr_2021)
# Make NULL variables with repetitive data
hr_2021[13:20] <- NULL
hr_2021[4:6] <-NULL
```

# Exploratory Data analysis 
## Correlation 
### Scatter-plot matrix

```{r, message=FALSE}
#Scatter-plot matrix
ggpairs(hr_2021[3:9])
#heat map
#Positive correlations are displayed in blue and negative correlations in red.
#Color intensity and the size of the circle are proportional to the correlation coefficients.
corrplot(cor(hr_2021[3:9]), method = "color") 
```

With "ggpairs" from the libraries "ggplot2" and "GGally", we visualize   
the different correlations between variables in a scatter plot matrix, better than 
the usual matrix that we obtain with "plot", as it provides one score, and proper  
visualization of the distribution of data in each of the variables. 
We can see that the variables about GDP, social support, and healthy life expectancy, 
have the highest correlation with our dependent variables "Ladder.score". Meaning that has these variables 
increase, so does the ladder score. We can still see correlation between the variables, 
freedom to make life choices, generosity, and perception of corruption, but they 
are not as strong as the other three variables mentioned before.
Our heat map with "corrplot" function, from the package "corrplot", supports our 
previous observations, as positive correlations are displayed in blue and negative 
correlations in red. Th color intensity and the size of the 
are proportional to the correlation coefficients.

# Regression Models

Our goal with our regression analysis, will be to predict Ladder.score . We will 
perform different models, using singular, multiple and all the variables. 

Our exploratory data analysis showed us the variables that are more correlated 
with Ladder.score. 

Model 1 will be a simple linear regression will be a simple linear regression L
adder.score ~ Logged.GDP.per.capita. 

Model 2 will contain all the variables that showed more correlation with Ladder.score, 
which are: Logged.GDP.per.capita, Social.support, Healthy.life.expectancy and 
Freedom.to.make.life.choices.  

Model 3 will have all the variables in the dataset. Including interaction between 
the nominal variable of region. 

We will use the summary() function to obtain an analysis of each model, which will 
be of use when selecting the best one.  

```{r}
#Model 1 
model_1 <- lm(Ladder.score  ~ Logged.GDP.per.capita , data = hr_2021)
##Model 1 summary 
summary(model_1)

#Model 2
model_2 <- lm(Ladder.score ~ Logged.GDP.per.capita + Social.support + 
                Healthy.life.expectancy + Freedom.to.make.life.choices, data = hr_2021)
##Model 2 summary 
summary(model_2)

#Model 3
#Data used for model 3, since we dont want name of countries on the linear regression
model_3_data <- hr_2021[3:9]
model_3 <- lm(Ladder.score ~ ., data = model_3_data)
##Model 3 summary 
summary(model_3)
```

We will select model 2, since it explains variance of 74.42% compared to model 1
which explains 62.37%, and compared to model 3, which explains variance of 75.58%,
that is higher than our model 2, howevere model 2 has only 4 variables while 
model 3 has 7.

```{r}
summary(model_2)
```

Summary of model 2 provide us with insights about the performance of the model,
such intercept and the coefficients of the independent variables and the 
dependent variable. 
A unit increase in Logged.GDP.per.capita will increase the score by 0.292 in
Ladder.score.
Summary of model 2 also shows us  that all the 4 variables are statistically significant, 
with values lower that 0.05.
In addition R square of 74.42% indicates that the variation is explained by the model is good. 
For now, we will move to try to improve the results of the summary of model 2.

# Diagnostic and transformations
## Regression diagnostic plots

```{r}
#Regression diagnostics on model 2
par(mfrow = c(2,2)) 
plot(model_2)
```

 - Residual vs Fitted: Doesn't show a great horizontal line, looks more like a  
U shape. At least there is not a distinct pattern which is an indicator for relationship. 

- Normal Q-Q: Good residual distribution as the points follow the straight dashed line. 
just a few outliers on the lower left side, and the upper right side. 

- Scale-Location (or Spread-Location): Shows some variance in our residuals,  
as we don’t have a horizontal line, but some inclination. 

- Residuals vs Leverage: Shows there are few extreme values that would affect our regression,  
However, nothing too severe. 

## Standarized residuals against each predictor.

```{r}
par(mfrow = c(2,2)) 
rstand <- rstandard(model_2) 
plot(rstand ~ Logged.GDP.per.capita, data = hr_2021) 
plot(rstand ~ Social.support, data = hr_2021) 
plot(rstand ~ Healthy.life.expectancy, data = hr_2021) 
plot(rstand ~ Freedom.to.make.life.choices, data = hr_2021)
```

The standardized residuals plot shows Homoscedasticity, among the variables,  
since there are no cluster of data points. This plot also strength, our initial visualization 
of the correlation of the data, where we could see that most of the data showed normal 
distribution, good histogram. Implying that even transformations such as applying  
log() on our linear regression mode, will not have a significant change in the results  
previous results of our model. 

# Transformation
## Developing model based on Log transformations

```{r}
#log transformation of model 1
log_model_1 <- lm(log(Ladder.score) ~ Logged.GDP.per.capita + Social.support + 
    Healthy.life.expectancy + Freedom.to.make.life.choices, data = hr_2021)

summary(log_model_1)$r.sq

#log transformation on model 2 
log_model_2 <- lm(log(Ladder.score) ~ Logged.GDP.per.capita + Social.support * 
                    Freedom.to.make.life.choices + Healthy.life.expectancy, data = hr_2021)

summary(log_model_2)$r.sq
```

As expected, since the data on the World Happiness Report is on a good part, very  
good distributed, utilizing log() on our dependent variable, did not improve the  
results of our previous model 2, even after adding interaction between two of the  
variables that have a significant impact on our dependent variable (GDP and Social Support). 

# Conclusions

It seems out model 2 is the best fit, the summary also supports this, since the  
model has p-value less than 0.05, all the variables are statistically significant,  
low residual to standard error ratio, and well distributed data, as well, both 
previous log() on lm() didn't surpass the 74.42% obtained previously. 

We can see which variables are related the most to the "ladder.score (happiness score)" 
and it is remarkably interesting to see that happiness cannot be measured just by money  
(GDP per capita), or only for freedom. However, our model shows us that freedom of 
choice and life expectancy, has a significant impact on the happiness score. Of course, 
a combination on high scores on the other variables is important, but these two are
the ones with the biggest impact on the score. 

## Predictions based on selected model 
Our fitted regression model is the next:
Ladder.Score = -3.11157 + 0.29182(Logged.GDP.per.capita) – 2.16586(Social.support) 
+ 0.03308(Healthy.life.expectancy) + 2.49817(Freedom.to.make.life.choices)

Prediction case 1 - Balanced GDP - Freedom and life expectancy
Country with scores of: GDP = 9.750, Social support = 0.854, healthy life 
expectancy = 70.000, freedom to make life choices  = 0.9000

```{r}
#define new country
new <- data.frame(Logged.GDP.per.capita=c(9.775), Social.support=c(0.854),
                  Healthy.life.expectancy=c(70.000), Freedom.to.make.life.choices = 0.900)

#use the fitted model to predict the rating for the new player
predict(model_2, newdata=new)
```

Prediction case 2 - High GDP and low life expectancy and freedom
Country with scores of: GDP = 6.750, Social support = 0.8000, healthy life 
expectancy = 65, freedom to make life choices  = 0.8000

```{r}
#define new country
new <- data.frame(Logged.GDP.per.capita=c(11.500), Social.support=c(0.8000),
                  Healthy.life.expectancy=c(50), Freedom.to.make.life.choices = 0.4000)

#use the fitted model to predict the rating for the new player
predict(model_2, newdata=new)
```

Prediction case 3 - Low High GDP and high life expectancy and freedom
Country with scores of: GDP = 6.750, Social support = 0.8000, healthy life 
expectancy = 65, freedom to make life choices  = 0.8000

```{r}
#define new country
new <- data.frame(Logged.GDP.per.capita=c(6.635), Social.support=c(0.983),
                  Healthy.life.expectancy=c(76.953), Freedom.to.make.life.choices = 0.970)

#use the fitted model to predict the rating for the new player
predict(model_2, newdata=new)
```

## Limitation and future improvements
There are improvements that could be made to the analysis, the data was well distributed overall, 
however, for a future project, we could gather data from previous year, and that 
will improve our prediction power, as all models feed of the amount of data available. 
We could also do comparison on year that there were some catastrophes or big event 
for the world and see if the independent variables correlation with the happiness 
score changed.  

## Thank you 


